---
title: "Markdown Reports of Data Analyses -- BRFSS example"
author: "Lianne Sheppard"
date: "Monday, February 22, 2016"
output: 
  ioslides_presentation:
        fig_width: 8.5
        fig_height: 3.5
        keep_md: yes
        smaller: yes 
  html_document:
        fig_caption: yes
        
---
# Writing Reports in Rmarkdown

---  

## Outline
* Learning objectives
* Rmarkdown basics
* Report outline and structure  
* Putting it all together:  BRFSS example  

---

## Learning objectives
* Become familiar with Rmarkdown  
* Understand the role of good reports in reproducible research    

---  

## Why am I focusing on this?
* Just as we should always approach statistical analyses with a plan in mind, it is very helpful to have a structured approach to reporting the full scope of our analyses
* We want to document our work to support it being reproducible
* We should summarize our work in a fashion that is useful to all collaborators
     * This means helping readers glean the important results from the first few pages of the report   

---  

## R Markdown (.Rmd) basics  

Format for writing reproducible, dynamic reports with R. A core tool in literate statistical programming.

Embed R code and results into slideshows, pdfs, and html documents.

Allows the user to follow a simple approach -- easy to type and read.  Based on plain text with certain text features that tell R and Markdown how to treat the text. Works with any basic text editor.

Easy to manage workflow in Rstudio.

See the R Markdown [cheatsheet](http://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)

--- 

## Formatting  

#### Use simple typing and a few symbols to get most of whay you want:  

Plain text stays the same

Make use of empty lines and spaces at the end of lines to start new paragraphs or lines   


```
Use *italics* or _italics_ 
and **bold** or __bold__
```

Use *italics* or _italics_ 
and **bold** or __bold__

--------------------------

```
superscripts^2^
```
superscripts^2^  

---  

## Headers:  

```
Use # to indicate headings, one # for every level:  
# Header Level 1
## Header Level 2
### Header Level 3, etc

```
Use # to indicate headings, one # for every level:  

# Header Level 1

## Header Level 2

### Header Level 3, etc

---  

## Lists:  
```
* Item 1  
* Item 2  
  * Sub-item 1 -- indent enough to indicate  
  + Sub-item 2  -- tab or spaces 
1. Ordered item 1  
2. Ordered item 2 
4. Ordered item 3 -- note the numbers you assign don't matter
3. Ordered item 4
```

* Item 1  
* Item 2  
    * Sub-item 1 -- indent enough to indicate 
    + Sub-item 2 -- tab or spaces   
1. Ordered item 1  
2. Ordered item 2 
4. Ordered item 3 -- note the numbers you assign don't matter
3. Ordered item 4

---

## Embed R code

**Inline code** -- Surround code with back ticks and r:

```
Two plus two equals `r 2 + 2`
```
Two plus two equals `r 2 + 2`

**Code chunks** -- Start a chunk with 3 back ticks and {r}.  Insert a descriptive name after r in order to name the code chunk.  End the chunk with 3 back ticks:

<pre>
Sample code:
```&#123;r&#125;
dim(iris)
```
</pre>

Sample code:
```{r SampleCodeChunk}
dim(iris)
```

Use display options (eval, echo) to tell R whether to evaluate or show code

---   

## Report outline  
* Executive summary (text only; max 2 pages)
    + Introduction
    + Data & methods
    + Results
    + Conclusions
* Loading libraries and data
    * Libraries
    * Data
* Full data analysis (up to 15 pages)
* Appendix

---

## Let's start our Rmarkdown report

1. Open Rstudio
2. New file -- Rmarkdown
3. Choose a title, etc **ADD**
4. Insert some headers and section breaks:

```
---

# Executive summary

## Introduction

## Data & methods

## Results

## Conclusions

# Loading libraries and data

## Libraries

## Data

# Full data analysis

# Appendix

```

5. Save and Knit

---

## What should a report contain?

1. Clearly state the underlying scientific question you hope to address
2. Include a description of the experiments or observational study design with mention of how these are expected to answer the question (_a priori_ expectations about the results)
      * If this report is a subset of the full project, note what part of the analysis is in this report
3. Briefly summarize the results of the analysis
4. Describe the conclusions to be drawn from the results and next steps
5. At the end present the full analysis
6. Wrap up with an appendix that includes at least a few administrative details

---

## Executive summary

Write in prose, ideally <= 2 pages.
Should be clear to everyone on the project team.  

#### Outline

1. Introduction
    * Background
    * Rationale
2. Data & methods
    * Data description
    * Statistical analysis approach
3. Results
4. Conclusions

---

## Introduction

* Background and rationale:
    * Why are we doing this?
    * Who are we doing it for?  
    * Why is this a reasonabl thing to try?  

* Objectives
    * What are we trying to establish?
    * What outcomes would constitute success?

---

## Data & methods

##### What are our data?  
Where did our data come from? (Supply URLs as appropriate)
When did we get the data?
What are we measuring?
How many samples/subjects?  
How many measurements per subject/sample?  
What covariates are important?  

##### How are we processing the data?  
If the data will be filtered or grouped, what cutoffs will be used?  

##### Describe the analysis approach.  
What statistical tests are being used?
What sensitivity analyses are planned?



---

## Results

Use this section to provide an *objective summary* of the findings; interpretation goes in the conclusions section.

What did our statistical analyses show?
    * Reference the figures and tables that show this  (Don't include them in the executive summary)
    
Do our results pass sanity checks?

What summary files accompany this analysis?
    * List files produced by the analysis by name.

Anticipate questions from collaborators and answer them.

---

## Conclusions

The concluions provide the context for the results.  This is the place for interpretation in light of our objectives.

For each objective, include a statement or two about how that objective has or hasn't been met.
    * Keep reporting of objectives in the same order.
    * Address:  Given the data did our initial hypothesis seem plausible?

This section can include a discussion of the implications of the findings.  Also include data quality issues, caveats or limitations of the approach, and possible next steps.

---

## Loading libraries and data

For the libraries section, include text such as:

```{r loadLibraries}

library(knitr)
library(ggplot2)

```

For the data subsection, include both text to describe and code to access the data.  All steps for reading and preprocessing the data belong here.

---

## Full data analysis

This section should include: 

* Descriptive data analyses and verification that you are analyzing the data you intend to analyze

* Inferential analyses

* Important figures and tables  

---

# Appendices

Include any additional details that don't belong in the full data analysis but need to be included.

Also put in some documentation about the report, updated every time the report is run:
```
> sessionInfo()
> getwd()
```

For instance:
```{r getInfo, echo=TRUE}

getwd()
sessionInfo()

```

---

# What can we write before analyzing data? 

A surprising large amount of the report can be written before the detailed analysis has begun.

You can use the `eval=FALSE` option to prevent the code from being run before you are ready

Circulating drafts of the introduction plus data & methods sections before the data analysis begins can be very helpful and *save time*

---

# Strategies for incorporating R code

* Precede each block of code with a statement of the purpose of this code block

* Try not to include blocks of code that are longer than 1/3 of a page.  

* If a block of code processes data, include a few lines of data before and after to show how the processing worked.  (e.g., Use `head()` and `tail()`).

* Write functions for blocks of code that are reused extensively.
      * Use descriptive function names  
      
* Use descriptive names for datasets and variables
     * Name data frames and matices as well as their columns (and their rows, as appropriate)
     * Refer to your entries by name, not number


---

# Tips for clarity

* Does everyone on your team have a common understanding of the goals?

* Make sure it is clear what negative and positive mean in a each specific context

* Think about what is inferred from each plot you show
    * Describe what allows you to make this inference
    * Think about what the plot would look like if there were no structure in the data  
    
* Make sure it is clear what each chunk of code is meant to do

---

# Some ideas for sanity checks

* How do you know you are analyzing the data you think you are analyzing?  
     * Have you verified the sample size is correct at each data massaging step?
     * How do you know data have been merged correctly?
     
* Have you clearly described the data?  
    * What low-dimensional summaries of the data have you plotted?  Do they make sense?
    

* What do you expect to see if the data are purely noise?

---

## Writing a report of a data analysis

Our goal is to replicate the analysis of sleep recently published in the MMWR (ADD DETAILS) and produce their chloropleth map.

We will develop an R Markdown report of our work.  We will use the R code in the file healthy_sleeping.R for the code in our R Markdown report.  

Start by cloning from github:  <https://github.com/liannesheppard/brfssReport.git>

Take the code from the R file and insert it as appropriate into your R Markdown report.  Also insert text in the appropriate report sections.  Hand in a compiled version of your report.





