---
title: "Report Template"
author: "Keith Baggerly"
date: "November 19, 2015"
output:
  html_document:
    fig_caption: yes
    highlight: pygments
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 3
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 3
  word_document:
    fig_caption: yes
    highlight: tango
---

# Executive Summary

## Introduction

Who is this for?  
What is the motivating question?  
What is the objective of this report?  

## Data and Methods

Where is the data?  
When did we get it?  
What are we measuring?  
How many samples do we have?  
What groups are we contrasting?  
What statistical tests will we use?  
How will we account for multiple testing?  

Are there some differences we expect to see _a priori_? (positive controls)  
Do we know what "no structure" should look like? (negative controls)  

## Results

What did our statistical tests show?  
What figures illustrate this? How?  
Do our results pass simple sanity checks?  
What summary files (if any) accompany this report?  

## Conclusions

Given the data, does our initial hypothesis seem plausible?  

# Loading Libraries and Data

## Libraries

Here we load all of the R libraries we'll be using in this analysis.

```{r loadLibraries}

library(knitr)
library(knitrBootstrap)

```

## Data

# Analysis

```{r someCodeExample}

x <- 5
y <- 10
x + y

```

```{r someFigureExample}

plot(c(1:10), c(1:10)^3, xlab="X", ylab="X^3")

```


# Appendix

```{r getInfo, echo=TRUE}

getwd()
sessionInfo()

```

